Contracts for Contracts

### 2023-06-14 paper/contract_solidity.tex
sec 2

\subsection{Flat Contracts for First-Order Values}

@@

So far, boring.

Behavioral contracts for first-order values are straightforward,
and can be directly mapped to \code{assert} or \code{require} statements
appearing at the beginning of the function.

\subsection{Contracts for Higher-Order Functions}

\iffalse
	\begin{lstlisting}
function map(uint[] memory data, function (uint) pure returns (uint) f)
  internal pure returns (uint[] memory r)
{
  r = new uint[](data.length);
  for (uint i = 0; i < data.length; i++) {
    r[i] = f(data[i]);
  }
}
\end{lstlisting}
\fi

Function contracts are first-class -- so if this guarded function is escaped
(e.g. by returning), the contract is still enforced:
\begin{lstlisting}
NEED AN EXAMPLE
\end{lstlisting}

\subsection{Contracts for Addresses}

Address contracts share similar syntax as function contracts, but introduces
additionally more binders for the arguments and returned values:

\begin{lstlisting}
  # { a | { {value, gas, ...} | <pre-cond-value-gas> }
          -> { arg | <pre-cond-arg> }
          -> { (res, data) | <post-cond> } }
  function f(address a) {
    (bool success, bytes memory data) = a.call{value: ...}(arg);
  }
\end{lstlisting}

It would be convenient to directly enforces the success of the call and omit
the data:
\begin{lstlisting}
  # { a | { {value, gas}(arg) | <pre-cond> } -> { (true, _) | true } }
  function f(address a) { ... }
\end{lstlisting}

Address call can designate the callee function, therefore it is necessary to
specify the spec for possible callees.
\zz{just FYI, `a.f` can also specify `value`, `gas`, etc.}

@@

\wac{We may need an example here since the necessity to designate the callee function may not be obvious to reviewers.}
\zz{I guess we should educate the audience about the difference between low-level and high-level calls at the beginning of Section 2.4. I'll take care of this later.}

\subsubsection*{Limitations}
Address contracts are \emph{not} higher-order -- if an address takes another
address as argument, then we cannot enforce the contract of the argument
address.
For example, in the following case, \code{b} itself has an ``address contract''
\code{<b-pre> -> <b-post>} but there is no general way to enforce that.
Because we do not have the control of the actual callee function of \code{a},
therefore no way to modify the code of \code{a} to check the pre-condition
and post-condition when \code{b} is called in \code{a}.

@@

\subsection{Temporal Properties}

>>>>

\subsection{Contracts for First-Order Values}

@@

So far, these flat contracts for first-order values are straightforward,
and can be directly mapped to assertions wrapping around functions.

\subsection{Contracts for Higher-Order Values}

\paragraph{Addresses}
Resembling pointers in C/C++, addresses in Solidity are numerical values
that can be compared or \note{computed?}.
In this sense, specifications of addresses are no different
from other integer values.
For instance, the following specification requires that
the address argument value cannot be 0:
\begin{lstlisting}
# { f(addr) requires addr == 0 }
function f(address payable addr) { ... }
\end{lstlisting}

However, addresses in Solidity have rich higher-order behaviors: they may
represent a deployed Ethereum contract and can be invoked in customized ways.
For example, the following snippet shows a function \code{f} taking an address
\code{addr} that is called at some point.
The programmer intends to the specify the precondition and postcondition when
calling \code{addr}. \note{GW: example abi?}
Nevertheless, such checks cannot be performed when \code{f} is called,
even though \code{addr} is a proper argument of \code{f}.
Because \code{addr} in this use case is effectively a function, and
it is undecidable to statically examine its properties in general.
Additionally, we may not have access to the source code of the contract
represented by \code{addr}.
\begin{lstlisting}
# { f(addr) where
    { addr(msg, amount) returns (flag, data)
      requires amount > 100
      ensures flag == true }
function f(address payable addr) {
  ... // omitting statements before the call
  (bool success, bytes memory data) =
    addr.call(abi.encodeWithSignature(
        "foo(string, uint256)", "call foo", 128
      )
    );
  ... // omitting statements after the call
}
\end{lstlisting}
\lang addresses this issue by adopting lazy checks
\cite{DBLP:conf/icfp/FindlerF02} that only take place
when the address is actually invoked \note{GW: fwd ref to impl}.
The above snippet specifies the condition of \code{addr}
in the \code{where}-clause, which requires
the \code{amount} argument must be greater than 100 and
the address call must succeed.

\paragraph{Additional Arguments}
Solidity address calls can take additional special arguments
such as \code{value} and \code{gas}:
\begin{lstlisting}
addr.call{value: msg.value, gas: 5000}(...);
\end{lstlisting}
In \lang, programmers can specify their conditions by introducing additional
bindings using the familiar syntax:
\begin{lstlisting}
# addr{value: v, gas: g}(msg, amount) returns (flag, data)
\end{lstlisting}
\code{v} and \code{g} represent the actual message value and gas
value, which can be used in \code{requires}/\code{ensures}-clauses.

\paragraph{Multiple Targets}
Multiple callee targets can inhabit in a single address value.
Programmers can specify different conditions for different
callee targets in the \code{where}-clause.
\begin{lstlisting}
TODO EXAMPLE
\end{lstlisting}

\zz{I guess we should educate the audience about the difference between low-level and high-level calls at the beginning of Section 2.4. I'll take care of this later.}

\iffalse
It would be convenient to directly enforces the success of the call and omit
the data:
\begin{lstlisting}
  # { a | { {value, gas}(arg) | <pre-cond> } -> { (true, _) | true } }
  function f(address a) { ... }
\end{lstlisting}

\zz{just FYI, `a.f` can also specify `value`, `gas`, etc.}

@@

\wac{We may need an example here since the necessity to designate the callee function may not be obvious to reviewers.}
\fi

\paragraph{Second-Class Enforcement}
\note{GW: an easy option is to discard enforcement of escaped addresses at all.
But it seems we can guard escaped addresses within the current contract scope.
We need to think about layered/stacked contracts of addresses (lax/picky).}
Address contracts are \emph{not} higher-order -- if an address takes another
address as argument, then we cannot enforce the contract of the argument
address.
For example, in the following case, \code{b} itself has an ``address contract''
\code{<b-pre> -> <b-post>} but there is no general way to enforce that.
Because we do not have the control of the actual callee function of \code{a},
therefore no way to modify the code of \code{a} to check the pre-condition
and post-condition when \code{b} is called in \code{a}.

@@

\paragraph{Notes on Higher-Order Functions}
Functions in Solidity are first-class too, meaning that functions can be used
as arguments for and returned from functions. However, at this moment
\footnote{Solidity version 8.20} Solidity has not supported writing anonymous
functions (lambda expressions) or nested, named functions.
Only explicitly defined top-level functions can be used in first-class ways,
which effectively do not have capturing from the environment.
This restriction discourages programmers to use higher-order functions due to
its inconvenience and verbosity.
Moreover, it poses challenges to seamlessly implement first-class behavioral
contract and monitor system (e.g. as in \cite{DBLP:conf/icfp/FindlerF02})
without an expensive whole program transformation.
Therefore, in this paper we focus on the contract and monitoring of addresses
values and leave monitoring for higher-order functions as future work when
Solidity has a proper support for lambda expressions.

\iffalse
	\begin{lstlisting}
function map(uint[] memory data, function (uint) pure returns (uint) f)
  internal pure returns (uint[] memory r)
{
  r = new uint[](data.length);
  for (uint i = 0; i < data.length; i++) {
    r[i] = f(data[i]);
  }
}
\end{lstlisting}

Function contracts are first-class -- so if this guarded function is escaped
(e.g. by returning), the contract is still enforced:
\begin{lstlisting}
NEED AN EXAMPLE
\end{lstlisting}
\fi

\subsection{Contracts for Temporal Behaviors}
